%
% Copyright (c) 2007 Rathaxes Team (team@rathaxes.eu)
% 
% Permission to use, copy, modify, and distribute this software for any
% purpose with or without fee is hereby granted, provided that the above
% copyright notice and this permission notice appear in all copies.
% 
% THE ARTICLE IS PROVIDED "AS IS" AND THE AUTHOR DISCLAIMS ALL WARRANTIES
% WITH REGARD TO THIS ARTICLE INCLUDING ALL IMPLIED WARRANTIES OF
% MERCHANTABILITY AND FITNESS. IN NO EVENT SHALL THE AUTHOR BE LIABLE FOR
% ANY SPECIAL, DIRECT, INDIRECT, OR CONSEQUENTIAL DAMAGES OR ANY DAMAGES
% WHATSOEVER RESULTING FROM LOSS OF USE, DATA OR PROFITS, WHETHER IN AN
% ACTION OF CONTRACT, NEGLIGENCE OR OTHER TORTUOUS ACTION, ARISING OUT OF
% OR IN CONNECTION WITH THE USE OR PERFORMANCE OF THIS SOFTWARE.
%

\documentclass[11pt]{report}
\usepackage{listings}
\lstset{language=C}
\lstloadlanguages{C}
\lstset{numbers=left, breaklines=true, basicstyle=\ttfamily,
  numberstyle=\tiny\ttfamily, framexleftmargin=6mm,
  backgroundcolor=\color{grey}, xleftmargin=6mm, language=C,
  showspaces=false, showstringspaces=false}
\usepackage{color}
\usepackage[english]{babel}
\usepackage{listings}
\definecolor{grey}{rgb}{0.95,0.95,0.95}

\begin{document}
\author{Amsallem David\\Dumont Mickael\\Gallon Sylvestre\\Suarez Tomas}
\title{IO Access}
\maketitle

\section*{Licence}
\addcontentsline{toc}{section}{Licence}
Copyright (c) 2007 Rathaxes Team (team@rathaxes.eu)
\\\\
Permission to use, copy, modify, and distribute this software for any
purpose with or without fee is hereby granted, provided that the above
copyright notice and this permission notice appear in all copies.
\\\\
THE ARTICLE IS PROVIDED "AS IS" AND THE AUTHOR DISCLAIMS ALL WARRANTIES
WITH REGARD TO THIS ARTICLE INCLUDING ALL IMPLIED WARRANTIES OF
MERCHANTABILITY AND FITNESS. IN NO EVENT SHALL THE AUTHOR BE LIABLE FOR
ANY SPECIAL, DIRECT, INDIRECT, OR CONSEQUENTIAL DAMAGES OR ANY DAMAGES
WHATSOEVER RESULTING FROM LOSS OF USE, DATA OR PROFITS, WHETHER IN AN
ACTION OF CONTRACT, NEGLIGENCE OR OTHER TORTUOUS ACTION, ARISING OUT OF
OR IN CONNECTION WITH THE USE OR PERFORMANCE OF THIS SOFTWARE.
\newpage

\section*{Introduction}
\addcontentsline{toc}{section}{Introduction}

Rathaxes Team is divided into three work teams:\\
\begin{description}
  \item{The Transverse Team, in charge of the logistic part of the project and
    relaying informations between the two other teams.}
  \item{The Hardware Team, in charge of the system and low level aspects of the
    project and providing the Language Team with informations they need.}
  \item{The Language Team, in charge of designing and implementing the Rathaxes
    DSL, according to informations provided by the Hardware Team.\\}
\end{description}

As agreed within the Rathaxes teams, Rathaxes will be able to generate drivers
for the four following operating systems:
\begin{description}
    \item{OpenBSD}
    \item{Linux 2.6}
    \item{MacOS X 10.5}
    \item{Windows XP}
\end{description}

This paper deals with the concept of IO Access. It provides
an abstraction layer to describe the API of the different kernel
to access on device registers.

% distinction des differentes familles

\tableofcontents

\chapter{IO Access}

\section{IO ports}
Polling stuff;

\section{Mapped Memory}
Queue stuff.

\section{others}
Other stuff.

\section{others}
Other stuff.

\section{others}
Other stuff.

\end{document}
